---
title: "Assignment-4"
author: "Tushar Sharma"
format: pdf
editor: visual
---

```{r}
library(ggplot2)
```

```{r}
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ,y=hwy))
```

```{r}
ggplot(data = mpg) +
geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

```{r}
ggplot(data = mpg) +
geom_point(mapping = aes(x = displ, y = hwy, size = class))
```

```{r}
# Top
ggplot(data = mpg) +
geom_point(mapping = aes(x = displ, y = hwy, alpha = class))

```

```{r}
# Bottom
ggplot(data = mpg) +
geom_point(mapping = aes(x = displ, y = hwy, shape = class))

```

#### Can also set the aesthetic properties of your geom manually.

```{r}
ggplot(data = mpg) +
geom_point(mapping = aes(x = displ, y = hwy), color = "green",shape=23)
```

## Facets

-   useful for categorical variables

-   Is to split your plot into facets

-   subplots that each display one subset of the data.

```{r}
ggplot(data = mpg) +
geom_point(mapping = aes(x = displ, y = hwy)) +
facet_wrap(~ class, nrow = 2)
```

#### To facet your plot on the combination of two variables, add

facet_grid() to your plot call. The first argument of facet_grid() is also a formula.

```{r}
ggplot(data = mpg) +
geom_point(mapping = aes(x = displ, y = hwy)) +
facet_grid(drv ~ cyl)
```

## Geometric Objects

```{r}
ggplot(data = mpg) +
geom_point(mapping = aes(x = displ, y = hwy))
```

```{r}
ggplot(data = mpg) +
geom_smooth(mapping = aes(x = displ, y = hwy))
```

### Both together.

```{r}
ggplot(data = mpg) +
geom_point(mapping = aes(x = displ, y = hwy)) +
geom_smooth(mapping = aes(x = displ, y = hwy))
```

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
geom_point(mapping = aes(color = class , shape = class)) +
geom_smooth()
```

```{r}
ggplot(data = mpg) +
geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))

```

```{r}
ggplot(data = mpg) +
geom_smooth(
mapping = aes(x = displ, y = hwy, color = drv),
show.legend = FALSE
)
```

#### Our smooth line displays just a subset of the mpg dataset, the

subcompact cars. The local data argument in geom_smooth() over rides the global data argument in ggplot() for that layer only:

```{r}

library(dplyr)
```

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
 geom_point(mapping = aes(color = class)) +
 geom_smooth(
 data = filter(mpg, class == "subcompact"),
 se = FALSE
 )
```

#### Statistical Transformations

```{r}
ggplot(data = diamonds) +
geom_bar(mapping = aes(x = cut))
```

```{r}
demo <- tribble(
~a,      
~b,
"bar_1", 20,
"bar_2", 30,
"bar_3", 40
)
ggplot(data = demo) +
geom_bar(
mapping = aes(x = a, y = b), stat = "identity"
)
```

```{r}
ggplot(data = diamonds) +
stat_summary(
mapping = aes(x = cut, y = depth),
fun.ymin = min,
fun.ymax = max,
fun.y = median
)
```

```{r}
ggplot(data = diamonds) +
geom_bar(mapping = aes(x = cut, y = ..prop..))
ggplot(data = diamonds) +
geom_bar(
mapping = aes(x = cut, fill = color, y = ..prop..)
)
```

```{r}
#Position Adjustments
ggplot(data = diamonds) +
geom_bar(mapping = aes(x = cut, color = cut))
```

```{r}
ggplot(data = diamonds) +
geom_bar(mapping = aes(x = cut, fill = cut))
```

```{r}
#the bars are automatically stacked. Each colored rectangle represents a combination of cut and clarity:
ggplot(data = diamonds) +
geom_bar(mapping = aes(x = cut, fill = clarity))

```

```{r}
#position = "dodge" places overlapping objects directly beside one another.
ggplot(data = diamonds) +
geom_bar(
mapping = aes(x = cut, fill = clarity),
position = "dodge"
)
```

### Cordinate Systems

```{r}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) +
geom_boxplot()
```

```{r}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) +
geom_boxplot() +
coord_flip()
```

```{r}
bar <- ggplot(data = diamonds) +
geom_bar(
mapping = aes(x = cut, fill = cut),
show.legend = FALSE,
width = 1
) +
theme(aspect.ratio = 1) +
labs(x = NULL, y = NULL)

```

```{r}

bar + coord_flip()
```

```{r}
bar + coord_polar()
```

# Working on ***txhousing*** dataset

```{r}

library(dplyr)

data(txhousing)
head(txhousing)

```

```{r}
tx_clean <- na.omit(txhousing)
```

### trend over time

```{r}
#meadian house price overtime
ggplot(tx_clean, aes(x = date, y = median)) +
  stat_summary(fun = mean, geom = "line", color = "blue", linewidth = 1) +
  labs(
    title = "Average Median House Price in Texas Over Time",
    x = "Year",
    y = "Average Median Price"
  )

```

### Compare cites

-   each line represents a city
-   compare trends across cities

```{r}
top_cities <- tx_clean |>
  dplyr::count(city, sort = TRUE) |>
  head(6) |>
  dplyr::pull(city)

ggplot(
  dplyr::filter(tx_clean, city %in% top_cities),
  aes(x = date, y = sales, color = city)
) +
  geom_line(linewidth = 1) +
  labs(
    title = "Housing Sales Trend in Major Texas Cities",
    x = "Year",
    y = "Sales"
  )

```

### Histogram -Distribution of a numeric variable

```{r}
ggplot(tx_clean, aes(x = median)) +
  geom_histogram(binwidth = 10000, fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Median House Prices",
    x = "Median House Price",
    y = "Count"
  )

```

### Box Plot -meadian price by year

-   market variablity , shows how house price vary within each year.

```{r}
ggplot(tx_clean, aes(x = factor(year), y = median)) +
  geom_boxplot(fill = "skyblue") +
  labs(
    title = "Distribution of Median House Prices by Year",
    x = "Year",
    y = "Median House Price"
  )
```

### Sales by city

-compares sales variability acrooss big cities (visaulixe which cities are more stable )

```{r}
top_cities <- tx_clean %>%
  count(city, sort = TRUE) %>%
  head(6) %>%
  pull(city)

ggplot(filter(tx_clean, city %in% top_cities),
       aes(x = city, y = sales)) +
  geom_boxplot(fill = "lightgreen") +
  labs(
    title = "Sales Distribution in Major Texas Cities",
    x = "City",
    y = "Sales"
  )
```

### Inventory by Year -helps understand how housing supply changed year wise -high spread = unstable market

```{r}
ggplot(tx_clean, aes(x = factor(year), y = inventory)) +
  geom_boxplot(fill = "orange") +
  labs(
    title = "Housing Inventory Distribution by Year",
    x = "Year",
    y = "Inventory"
  )
```

### Average sale over time - high level trend of overall market activity

```{r}
ggplot(tx_clean, aes(x = date, y = sales)) +
  stat_summary(fun = mean, geom = "line", color = "blue", linewidth = 1) +
  labs(
    title = "Average Housing Sales Trend in Texas",
    x = "Year",
    y = "Average Sales"
  )
```

### Sales distribution -shows low or high sales for months

```{r}
ggplot(tx_clean, aes(x = sales)) +
  geom_histogram(fill = "steelblue", color = "black", bins = 30) +
  labs(
    title = "Distribution of Housing Sales",
    x = "Sales",
    y = "Count"
  )
```

### Mesian Price Distribution

-smooth view of price distribution -looks better than histogram for pattern recognition

```{r}
ggplot(tx_clean, aes(x = median)) +
  geom_density(fill = "purple", alpha = 0.4) +
  labs(
    title = "Density of Median House Prices",
    x = "Median Price"
  )
```

### scatter Plot

-   Show how pricing affects sales volume
-   over all relationship

```{r}
ggplot(tx_clean, aes(x = median, y = sales)) +
  geom_point(alpha = 0.4) +
  geom_smooth(se = FALSE, color = "red") +
  labs(
    title = "Relationship Between House Prices and Sales",
    x = "Median Price",
    y = "Sales"
  )
```

### sales trends by city

-each city has its own clean panel

```{r}

ggplot(filter(tx_clean, city %in% top_cities),
       aes(x = date, y = sales)) +
  geom_line(color = "darkgreen") +
  facet_wrap(~ city, scales = "free_y") +
  labs(
    title = "Sales Trends in Major Texas Cities",
    x = "Year",
    y = "Sales"
  )
```

### Average invetory by city -housing supply city-wise -helps in long terms avaibality

```{r}
ggplot(tx_clean, aes(x = city, y = inventory)) +
  stat_summary(fun = mean, geom = "bar", fill = "brown") +
  labs(
    title = "Average Housing Inventory by City",
    x = "City",
    y = "Average Inventory"
  ) +
  theme(axis.text.x = element_text(angle = 90))
```
