---
title: "BASIC COMMANDS (Assignment-3)"
author: "Tushar Sharma"
format: pdf
editor: visual
---

## Practicing Basic Operations of R

### Create a vector - c()

```{r}
#Vectors of number 
v1 <- c(1,2,3,4,5,6)
print(v1)

#vectorss of characters
v2 <- c("Anuj ", "Tushar", "datta ")
print(v2)

#missalaneous vector
v3 <- c(1,"hello", FALSE)
print(v3)
```

### Generating sequences of number - seq()

```{r}

s1 <- seq(1, 10)
print(s1)

s2 <- seq(1, 10, by = 2)
print(s2)

#equally spaced
s3 <- seq(1, 10, length.out = 5)
print(s3)
```

```{r}
### Length of vector - length()

l1 <- length(v1)
print(l1)

l2 <- length(v2)
print(l2)

l3 <- length(v3)
print(l3)
```

### Sorting values - sort()

```{r}
so1 <- sort(c(5, 2, 9, 1))
print(so1)

so2 <- sort(c(10, 4, 7), decreasing = TRUE)
print(so2)

so3 <- sort(c(3, 3, 1, 2))
print(so3)
```

### Ordering elements - order()

```{r}
o1 <- order(c(40, 10, 30))
print(o1)

o2 <- v1[order(v1)]
print(o2)

o3 <- v1[order(v1, decreasing = TRUE)]
print(o3)
```

### Reverse a vector - rev()

```{r}
r1 <- rev(v1)
print(r1)

r2 <- rev(v2)
print(r2)

r3 <- rev(c(TRUE, FALSE, TRUE))
print(r3)
```

### Unique values - unique()

```{r}
u1 <- unique(c(1, 1, 2, 3))
print(u1)

u2 <- unique(c("a", "b", "a"))
print(u2)

u3 <- unique(c(TRUE, TRUE, FALSE))
print(u3)
```

### Frequency table - table()

```{r}
t1 <- table(c(1, 1, 2, 3))
print(t1)

t2 <- table(v2)
print(t2)

t3 <- table(c("yes", "no", "yes"))
print(t3)
```

### Match values - match()

```{r}
m1 <- match(2, v1)
print(m1)

m2 <- match("Tushar", v2)
print(m2)

m3 <- match(10, v1)
print(m3)
```

### Membership test - %in%

```{r}
in1 <- 2 %in% v1
print(in1)

in2 <- "Anuj " %in% v2
print(in2)

in3 <- 100 %in% v1
print(in3)
```

### Find positions - which()

```{r}
w1 <- which(v1 > 3)
print(w1)

w2 <- which(v1 == 5)
print(w2)

w3 <- which(c(TRUE, FALSE, TRUE))
print(w3)
```

### Any condition TRUE - any()

```{r}
a1 <- any(v1 > 5)
print(a1)

a2 <- any(v1 < 0)
print(a2)

a3 <- any(is.na(c(1, NA, 3)))
print(a3)
```

### All conditions TRUE - all()

```{r}
al1 <- all(v1 > 0)
print(al1)

al2 <- all(v1 > 3)
print(al2)

al3 <- all(c(TRUE, TRUE, FALSE))
print(al3)
```

### Check missing values - is.na()

```{r}
na1 <- is.na(c(1, NA, 3))
print(na1)

na2 <- is.na(v1)
print(na2)

na3 <- is.na(c(NA, NA))
print(na3)
```

### Remove missing values - na.omit()

```{r}
no1 <- na.omit(c(1, NA, 3))
print(no1)

no2 <- na.omit(c(NA, 2, NA, 4))
print(no2)

no3 <- na.omit(c("a", NA, "b"))
print(no3)
```

### Replace values - replace()

```{r}
rep1 <- replace(v1, v1 > 4, 0)
print(rep1)

rep2 <- replace(v1, v1 == 3, 99)
print(rep2)

rep3 <- replace(c("a", "b", "c"), 2, "z")
print(rep3)
```

### Combine strings - paste()

```{r}
p1 <- paste("Hello", "World")
print(p1)

p2 <- paste("Data", "Science", sep = "-")
print(p2)

p3 <- paste(v2, collapse = ", ")
print(p3)
```

### Combine strings without space - paste0()

```{r}
p01 <- paste0("R", "Language")
print(p01)

p02 <- paste0("Roll", 1:3)
print(p02)

p03 <- paste0(v2, "Student")
print(p03)
```

### Convert to uppercase - toupper()

```{r}
up1 <- toupper("hello")
print(up1)

up2 <- toupper(v2)
print(up2)

up3 <- toupper(c("data", "science"))
print(up3)
```

### Convert to lowercase - tolower()

```{r}
lo1 <- tolower("HELLO")
print(lo1)

lo2 <- tolower(v2)
print(lo2)

lo3 <- tolower(c("DATA", "ML"))
print(lo3)
```

### Substring extraction - substr()

```{r}
sb1 <- substr("datascience", 1, 4)
print(sb1)

sb2 <- substr("machinelearning", 8, 15)
print(sb2)

sb3 <- substr(v2, 1, 3)
print(sb3)
```

### Repeat values - rep()

```{r}
rp1 <- rep(1, 5)
print(rp1)

rp2 <- rep(c(1,2), times = 3)
print(rp2)

rp3 <- rep(c("a","b"), each = 2)
print(rp3)
```

### Sum of values - sum()

```{r}
sm1 <- sum(v1)
print(sm1)

sm2 <- sum(c(1, 2, 3, NA), na.rm = TRUE)
print(sm2)

sm3 <- sum(1:10)
print(sm3)
```

### Mean of values - mean()

```{r}
mn1 <- mean(v1)
print(mn1)

mn2 <- mean(c(1, 2, 3, NA), na.rm = TRUE)
print(mn2)

mn3 <- mean(1:10)
print(mn3)
```

### Maximum value - max()

```{r}
mx1 <- max(v1)
print(mx1)

mx2 <- max(c(3, 7, 2))
print(mx2)

mx3 <- max(1:100)
print(mx3)
```

### Minimum value - min()

```{r}
mnm1 <- min(v1)
print(mnm1)

mnm2 <- min(c(3, 7, 2))
print(mnm2)

mnm3 <- min(1:100)
print(mnm3) 
```

## DATA MANIPULATION IN R

### Create a data frame - data.frame()

```{r}
# Simple data frame
df1 <- data.frame(
  id = c(1,2,3),
  name = c("A","B","C"),
  marks = c(70,85,90)
)
print(df1)

```

```{r}
# Data frame without factors
df2 <- data.frame(
  city = c("Delhi","Mumbai","Pune"),
  temp = c(35,30,28),
  stringsAsFactors = FALSE
)
print(df2)
```

```{r}
# Mixed data types
df3 <- data.frame(
  roll = 1:3,
  pass = c(TRUE, FALSE, TRUE)
)
print(df3)
```

### Structure and summary - str(), summary()

```{r}
str(df1)
summary(df1)
summary(df2)
```

### Access columns - \$, \[ \]

```{r}
df1$marks
df1[,2]
df1[, c("name","marks")]
```

### Add new columns

```{r}
df1$result <- c("Pass","Pass","Pass")
print(df1)

df1$grade <- ifelse(df1$marks >= 80, "A", "B")
print(df1)

df1$distinction <- df1$marks > 85
print(df1)
```

### Remove columns

```{r}
df1$result <- NULL
print(df1)

df1 <- df1[, -4]
print(df1)

df1 <- df1[, !(names(df1) %in% c("name"))]
print(df1)
```

### Filter rows - subset()

```{r}
subset(df1, marks >= 80)

```

### Conditional logic - ifelse()

```{r}
df1$status <- ifelse(df1$marks >= 75, "Pass", "Fail")
print(df1)

df1$grade2 <- ifelse(df1$marks >= 85, "A",
                     ifelse(df1$marks >= 70, "B", "C"))
print(df1)

ifelse(df1$marks > 80, TRUE, FALSE)
```

### Sorting data - order()

```{r}
df1[order(df1$marks), ]
df1[order(df1$marks, decreasing = TRUE), ]
df1[order(df1$grade, df1$marks), ]
```

### Aggregation - aggregate()

```{r}
aggregate(marks ~ status, data = df1, mean)
aggregate(id ~ status, data = df1, length)
aggregate(marks ~ status, data = df1, sum)

```

### Merge data frames - merge()

```{r}
df4 <- data.frame(
  id = c(1,2,3),
  age = c(20,21,22)
)

merge(df1, df4, by = "id")
merge(df1, df4, by = "id", all.x = TRUE)
merge(df1, df4, by = "id", all.y = TRUE)

```

### Apply family - apply(), lapply(), sapply()

```{r}
apply(df4[, "age", drop = FALSE], 1, mean)
apply(df4[, "age", drop = FALSE], 2, mean)
lapply(df4, class)
sapply(df4, class)

```

### Handling missing values

```{r}
vec1 <- c(1, NA, 3, NA)
is.na(vec1)
na.omit(vec1)
sum(vec1, na.rm = TRUE)
```

### Factors and levels

```{r}
df1$grade <- factor(df1$grade)
levels(df1$grade)
nlevels(df1$grade)
as.character(df1$grade)
```

### Rename columns

```{r}
colnames(df1)[colnames(df1) == "marks"] <- "score"
print(df1)
```

### Sample rows

```{r}
sample(df1$id)
sample(df1$id, 2)
sample(df1$id, replace = TRUE)
```

### Bind data frames - rbind(), cbind()

```{r}
df5 <- data.frame(
  id = c(4,5),
  score = c(60,95),
  grade = c("C","A"),
  distinction = c(FALSE, TRUE),
  status = c("Fail","Pass"),
  grade2 = c("C","A")
)

rbind(df1, df5)
cbind(df4, gender = c("M","F","M"))
```

### Basic transformations

```{r}
df1$score_scaled <- df1$score / max(df1$score)
print(df1)

log(df1$score)
sqrt(df1$score)
```

### Head and tail

```{r}
head(df1)
tail(df1)
```

## Control Flow & Programming

#### ifelse() – Vectorized if-else (works on each element)

```{r}
#   Check if numbers are even or odd
result1 <- ifelse(c(1, 2, 3, 4) %% 2 == 0, "even", "odd")
print(result1)

#   Mark values above 5 as "high", others "low"
result2 <- ifelse(c(3, 6, 9), "high", "low")
print(result2)
```

```{r}
#   Simple if-else statement
x <- 10
if (x > 5) {
  print("x is greater than 5")
} else {
  print("x is not greater than 5")
}

#   Using if-else inside a function
check_number <- function(num) {
  if (num < 0) {
    return("Negative")
  } else {
    return("Non-negative")
  }
}
print(check_number(-3))
print(check_number(4))
```

```{r}
#   Print numbers 1 to 5
for (i in 1:5) {
  print(i)
}

#   Loop over elements in a vector
vec_loop <- c("a", "b", "c")
for (letter in vec_loop) {
  print(letter)
}
```

```{r}
#   Count from 1 until 5
counter <- 1
while (counter <= 5) {
  print(counter)
  counter <- counter + 1
}

#   Decrease a value until it is less than 0
val <- 3
while (val >= 0) {
  print(val)
  val <- val - 1
}
```

```{r}
#   Use repeat loop to count until 3
count <- 1
repeat {
  print(count)
  count <- count + 1
  if (count > 3) break
}

#   Another repeat example with a condition
i <- 10
repeat {
  i <- i - 2
  print(i)
  if (i <= 4) break
}
```

```{r}
# Example: Use break inside a for loop when condition met
for (i in 1:10) {
  if (i == 6) {
    break  # exit loop when i is 6
  }
  print(i)
}
```

```{r}
# Example: Skip printing the number 3 in a loop
for (i in 1:5) {
  if (i == 3) {
    next  # skip the rest of this loop cycle
  }
  print(i)
}
```

```{r}
#   A simple function that adds two numbers
add_two <- function(a, b) {
  return(a + b)
}
print(add_two(3, 5))

#   Function with a default parameter
greet <- function(name = "friend") {
  paste("Hello", name)
}
print(greet("Alice"))
print(greet())
```

```{r}
#   A simple function that adds two numbers
add_two <- function(a, b) {
  return(a + b)
}
print(add_two(3, 5))

#   Function with a default parameter
greet <- function(name = "friend") {
  paste("Hello", name)
}
print(greet("Alice"))
print(greet())

# (Used above in the add_two and greet functions)
```

```{r}
#   Stop if an invalid condition is met inside a function
check_positive <- function(x) {
  if (x < 0) {
    stop("x must be non-negative!")
  }
  return(x)
}
# Uncomment the following line to see the error:
# check_positive(-5)

#   Stop execution manually
# Uncomment the next line to stop the script:
# stop("Manual stop here!")
```

```{r}
#   Issue a warning if a value is too high
warn_if_high <- function(x) {
  if (x > 100) {
    warning("x is very high!")
  }
  return(x)
}
print(warn_if_high(150))
print(warn_if_high(50))
```

```{r}
#   Try to take the square root of a negative number
result_try1 <- try(sqrt(-1))
print(result_try1)

#   Try reading a non-existent variable
result_try2 <- try(print(nonexistent_variable))
print(result_try2)
```

```{r}
#   Use tryCatch to handle an error
result_tc <- tryCatch({
  log("a")  # will cause an error
}, error = function(e) {
  "An error occurred"
})
print(result_tc)

#   tryCatch with finally block
result_tc2 <- tryCatch({
  stop("Oops!")
}, error = function(e) {
  "Caught the error"
}, finally = {
  cat("This is the cleanup code.\n")
})
print(result_tc2)
```

# Working with Files {#5-working-with-files}

#### getwd() -- Get current working directory

```{r}
#   Simply print the working directory
wd1 <- getwd()
print(wd1)

#   Save and print working directory from a variable
wd2 <- getwd()
cat("Working directory is:", wd2, "\n")
```

#### setwd() -- Set the working directory

```{r}
# (Be careful with this! It changes your folder for reading/writing files.)
#   Change to a temporary directory (this should work safely)
# setwd(tempdir())
# cat("Now in temporary directory:", getwd(), "\n")

#   Set back to original (if you saved it before)
# original_wd <- "your/original/path"  # replace with your actual path
# setwd(original_wd)
```

#### list.files() -- List files in a directory

```{r}
#   List files in current directory
files1 <- list.files()
print(files1)

#   List files with a specific pattern (e.g., files ending with .R)
files2 <- list.files(pattern = "\\.R$")
print(files2)
```

#### file.exists() -- Check if a file exists {#fileexists--check-if-a-file-exists}

```{r}
#   Check for a file named "test.txt"
exists1 <- file.exists("test.txt")
print(exists1)

#   Check multiple files at once
exists2 <- file.exists(c("test.txt", "nonexistent.txt"))
print(exists2)
```

#### file.remove() -- Delete a file

```{r}
#   Create and then remove a temporary file
temp_file <- tempfile(fileext = ".txt")
cat("Temporary content", file = temp_file)
removed1 <- file.remove(temp_file)
print(removed1)

#   Try removing a file that likely does not exist
removed2 <- file.remove("no_file.txt")
print(removed2)
```

#### read.csv() -- Read a CSV file

```{r}
#   Write a small CSV and read it back
df_sample <- data.frame(Name = c("A", "B"), Value = c(10, 20))
temp_csv <- tempfile(fileext = ".csv")
write.csv(df_sample, temp_csv, row.names = FALSE)
df_read1 <- read.csv(temp_csv)
print(df_read1)

#   Read CSV with header (default behavior)
df_read2 <- read.csv(temp_csv, header = TRUE)
print(df_read2)
```

#### write.csv() -- Write data to a CSV file

```{r}
#   Write a data frame to CSV
write.csv(df_sample, "sample_output.csv", row.names = TRUE)
# Check if file exists afterward
print(file.exists("sample_output.csv"))

#   Write CSV without row names
write.csv(df_sample, "sample_output2.csv", row.names = FALSE)
print(file.exists("sample_output2.csv"))
```

#### read.table() -- Read tabular data from a file

```{r}
#   Create a text file and read it as table
temp_txt <- tempfile(fileext = ".txt")
write("1 2 3\n4 5 6", file = temp_txt)
table1 <- read.table(temp_txt)
print(table1)

#   Read table with a specified separator (space here)
table2 <- read.table(temp_txt, sep = " ")
print(table2)
```

#### write.table() -- Write data to a file in table format

```{r}
#   Write a data frame to a text file
write.table(df_sample, "sample_table.txt")
print(file.exists("sample_table.txt"))

#   Write table with row names disabled
write.table(df_sample, "sample_table2.txt", row.names = FALSE)
print(file.exists("sample_table2.txt"))
```

#### saveRDS() -- Save a single R object to a file

```{r}
#   Save a data frame as an RDS file
saveRDS(df_sample, "sample_data.rds")
print(file.exists("sample_data.rds"))

#   Save a vector as an RDS file
saveRDS(c(1,2,3), "sample_vector.rds")
print(file.exists("sample_vector.rds"))
```

#### readRDS() -- Read a single R object from an RDS file

```{r}
#   Read the previously saved data frame
df_from_rds <- readRDS("sample_data.rds")
print(df_from_rds)

#   Read the vector from the RDS file
vec_from_rds <- readRDS("sample_vector.rds")
print(vec_from_rds)
```

# Working with Factors {#6-working-with-factors}

#### factor() -- Create a factor variable

```{r}
#   Convert a character vector to a factor
fact_a <- factor(c("low", "medium", "high", "medium"))
print(fact_a)

#   Specify levels manually
fact_b <- factor(c("red", "blue", "red"), levels = c("red", "blue", "green"))
print(fact_b)
```

#### levels() -- Get or set levels of a factor

```{r}
#   Get levels of a factor
levels_a <- levels(fact_a)
print(levels_a)

#   Change levels of a factor
levels(fact_a) <- c("L", "M", "H")
print(fact_a)
```

#### relevel() -- Change the reference level of a factor

```{r}
#   Make "blue" the reference level
fact_c <- factor(c("red", "blue", "red", "green"))
fact_c <- relevel(fact_c, ref = "blue")
print(fact_c)

#   Relevel with a different reference
fact_d <- factor(c("A", "B", "C", "A"))
fact_d <- relevel(fact_d, ref = "C")
print(fact_d)
```

#### cut() -- Divide numeric data into intervals (factors)

```{r}
#   Cut numbers into equal-width bins
nums <- c(1, 3, 5, 7, 9)
cut1 <- cut(nums, breaks = 3)
print(cut1)

#   Cut with custom labels
cut2 <- cut(nums, breaks = c(0, 4, 8, 10), labels = c("low", "medium", "high"))
print(cut2)
```

# Date & Time Handling {#7-date--time-handling}

#### as.POSIXct() -- Convert string to date-time

```{r}
#   Convert a date-time string
dt1 <- as.POSIXct("2025-02-03 12:34:56")
print(dt1)

#   Convert with a specific format
dt2 <- as.POSIXct("03/02/2025 12:34:56", format = "%d/%m/%Y %H:%M:%S")
print(dt2)
```

#### as.POSIXlt() -- Convert string to list-like date-time

```{r}
#   Convert and see the list components
dt_lt1 <- as.POSIXlt("2025-02-03 12:34:56")
print(dt_lt1)

#   Extract the year from the POSIXlt object
year_val <- dt_lt1$year + 1900  # year is stored as years since 1900
print(year_val)
```

#### difftime() -- Compute the difference between two date-time values

```{r}
#   Difference between two dates
diff1 <- difftime(as.Date("2025-02-10"), as.Date("2025-02-03"))
print(diff1)

#   Difference in hours between two date-times
dt_start <- as.POSIXct("2025-02-03 08:00:00")
dt_end   <- as.POSIXct("2025-02-03 20:00:00")
diff2 <- difftime(dt_end, dt_start, units = "hours")
print(diff2)
```

#### format() -- Format date-time objects as strings

```{r}
#   Format a date in a simple way
fmt1 <- format(as.Date("2025-02-03"), "%B %d, %Y")
print(fmt1)

#   Format a date-time to show time and date
fmt2 <- format(Sys.Date(), "%Y/%m/%d")
print(fmt2)
```

# Debugging & Environment Management

#### ls() -- List objects in the environment

```{r}
#   List all current objects
objs1 <- ls()
print(objs1)

#   List objects that start with "df"
objs2 <- ls(pattern = "^df")
print(objs2)
```

#### rm() -- Remove objects from the environment

```{r}
#   Remove a single variable
temp_var <- 123
print(temp_var)
rm(temp_var)
# Uncommenting the next line would cause an error since temp_var is removed:
# print(temp_var)

#   Remove multiple objects
a <- 1; b <- 2; c <- 3
rm(a, b)
# Only 'c' should remain
print(ls())
```

#### gc() -- Run garbage collection (free up memory)

```{r}
#   Run garbage collection and print result
gc_result1 <- gc()
print(gc_result1)

#   Just call gc() to see memory cleanup info
gc()
```

#### debug() -- Turn on debugging for a function

```{r}
#   Define a simple function and set debug on it
test_fun <- function(x) {
  y <- x + 1
  return(y)
}
debug(test_fun)
# Uncomment the next line to step through the function:
# test_fun(5)
undebug(test_fun)  # Turn off debugging
```

### traceback() -- Show the call stack after an error

```{r}
#   Force an error then show traceback
error_fun <- function() {
  stop("This is an error!")
}
# Uncomment the following lines to see traceback:
# try(error_fun())
# traceback()
```

#### print() -- Print output to the console

```{r}
#   Print a number
print(100)

#   Print a message
print("Hello, world!")

# Extra example: Print a vector
print(c("a", "b", "c"))
```

#### cat() -- Concatenate and print text (no quotes)

```{r}
#   Concatenate and print a simple message
cat("This is a simple message.\n")

#   Concatenate multiple pieces of text
cat("The answer is", 42, "\n")
# Extra example: Print without automatic newlines (you add \n manually)
cat("Line one.\nLine two.\n")
```
