---
title: "Assignment-6"
author: "Tushar Sharma"
format: pdf
editor: visual
---

## Reading a CSV File in R

```{r}
list.files()
fertility <- read.csv("bd-dec22-age-specific-fertility-rates.csv")
print(fertility)
```

### View Basic Structure

```{r}
head(fertility)
```

```{r}
tail(fertility)
```

```{r}
str(fertility)
```

```{r}
dim(fertility)
```

```{r}
print(is.data.frame(fertility))
print(nrow(fertility))
print(ncol(fertility))
```

### Summary Statistics

```{r}
summary(fertility)
```

#### Handling Missing Values

```{r}
sum(is.na(fertility))

#missing value per column
colSums(is.na(fertility))
```

## LearningLoading other data set also

```{r}
orders <- read.csv("orders_jan.csv")
print(orders)
```

```{r}
library(dplyr)

orders <- orders %>%
  mutate(Total_Sales = quantity * order_amount)

head(orders)
```

### Total reveneue

```{r}
sum(orders$order_amount, na.rm = TRUE)
```

### Top 20 highest value order

```{r}
orders %>%
  arrange(desc(order_amount)) %>%
  head(20)
```

### Total sales by Country

```{r}
orders %>%
  group_by(country) %>%
  summarise(Product_Sales = sum(order_amount, na.rm = TRUE)) %>%
  arrange(desc(Product_Sales))
```

### Total order by customer

```{r}
orders %>%
  count(customer_id) %>%
  arrange(desc(n))
```

### Average order value

```{r}
mean(orders$order_amount, na.rm = TRUE)
```

## Filtering Data

### Orders above 10000

```{r}
orders %>%
  filter(order_amount>10000)
```

### Orders for Specific Country

```{r}
orders %>%
  filter(country == "india")
```

## Sorting Data

```{r}
# Sort by Quantity descending
orders %>%
  arrange(desc(quantity))
```

## Vizualization

```{r}
library(ggplot2)

# Sales by country
ggplot(orders, aes(x = country, y = order_amount)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Quantity distribution
ggplot(orders, aes(x = quantity)) +
  geom_histogram(fill = "orange", bins = 10)
```

### Writing to New CSV

```{r}
# Save high value orders
high_value <- orders %>%
  filter(order_amount > 10000)

write.csv(high_value, "high_value_orders.csv", row.names = FALSE)
```

## Converting to Tibble

```{r}
library(tidyverse)

orders_tbl <- as_tibble(orders)

glimpse(orders_tbl)
```

## Using readr Instead of base R

```{r}
library(readr)

orders_readr <- read_csv("orders_jan.csv")

spec(orders_readr)
```

```{r}
library(purrr)

orders %>%
  select(where(is.numeric)) %>%
  map_dbl(mean, na.rm = TRUE)
```
