---
title: "Mini Project - Multicollinearity"
author: "G Radhesyam"
date: "1 January 2019"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#```{r  include=FALSE,echo=FALSE}
# Set Working Directory
setwd("/home/saisyam/Documents/ACADEMICS_TEACHING/MDH_Workshops/Jan_2019/MultiColliniarity")
getwd()
```
```{r include=FALSE}
#Load the libraries required.
library(MASS)
library(ggplot2) # plot
library(calibrate) # textxy??
library(car)#vif, durbin-watson test
library(orcutt)  # To fix atuto-correlation issue
library(leaps) #regsubset
library(qpcR)#To calculate PRESS Residual and PRESS Statistic
library(factoextra)## PCA graphs
library(nortest) #### for normality testing - Anderson-Darling etc
```

## Multicollinearity
A basic assumption is multiple linear regression model is that the rank of the matrix of observations on explanatory variables is same as the number of explanatory variables. In other words, such matrix is of full column rank. This in turn implies that all the explanatory variables are independent, i.e., there is no linear relationship among the explanatory variables. It is termed that the explanatory variables are orthogonal. 

In many situations in practice, the explanatory variables may not remain independent due to various reasons. The situation where the explanatory variables are highly intercorrelated is referred to as multicollinearity.

### Consequences of multicollinearity
To illustrate the consequences of presence of multicollinearity, consider a model

$y = β_1 x_1 + β_2 x_2 + ε , E ( ε ) = 0, V ( ε ) = σ^{2} I$

where $x_1$ , $x_2$ and y are scaled to length unity.

$Var ( b 1 ) = Var ( b 2 ) = \sigma^{2}/(1-r^{2})$

$Cov ( b 1 , b 2 ) = −r \sigma^{2}/(1-r^{2})$

where r is the correlation coefficient between $x_1$ and $x_2$

If $r → ± 1$,  then  $Var (b1) =Var (b2) → ∞$

So if variables are perfectly collinear, the variance of OLSEs becomes large. This indicates highly unreliable estimates and this is an inadmissible situation.

|     r            |    0.99        |    0.9       |     0.1          |    0       |
|------------------|----------------|--------------|------------------|------------|
| Var(b1)=Var(b2)  | $50\sigma^{2}$ |$5\sigma^{2}$ | $1.01\sigma^{2}$ |$\sigma^{2}$|

* If r is close to 0, then multicollinearity does not harm and it is termed as **non-harmful** **multicollinearity** .
* If r is close to +1 or -1 then multicollinearity inflates the variance and it rises terribly. This is termed as **harmful** **multicollinearity**.

## Hald Cement Dataset - Data Inspection

### Load HALD CEMENT Dataset   
```{r echo=FALSE}
haldcement <- read.csv("BEST-Hald Cement data.csv")
head(haldcement )
str(haldcement )
attach(haldcement)
```


### HALD CEMENT Dataset
* The response variable Y is the heat evolved in a cement mix.

* The four explanatory variables are ingredients of the mix,
    + X1:Tricalcium aluminate
    + X2:Tricalcium silicate,
    + X3:Tetracalcium alumino ferrite
    + X4:Dicalcium silicate.
    
* An important feature of these data is that the variables x1 and x3 are highly correlated (corr(x1,x3)=-0.824), as well as the variables x2 and x4 (with corr(x2,x4)=-0.975). Thus we should expect any subset of (x1,x2,x3,x4) that includes one variable from highly correlated pair to do as any subset that also includes the other member.
   

## MODEL ANALYSIS and DETAILS 
```{r echo=FALSE}
# So let us build an initial model with Y vs X1,X2,X3 and X4
HC_LM_1 <- lm(haldcement$Y ~ ., data = haldcement )

# What's the model - linear equation
#HC_LM_1

#Summary of the model ( R2 adjusted etc)
summary(HC_LM_1)
```
### Scatter plot  
```{r echo=FALSE}
library(ggplot2)
panel.cor <- function(x, y){
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- round(cor(x, y), digits=2)
  txt <- paste0("R = ", r)
  cex.cor <- 0.8/strwidth(txt)
  text(0.5, 0.5, txt, cex = cex.cor * r)
}
pairs(haldcement$Y ~ ., data = haldcement,lower.panel = panel.cor)
```

## Outlier Inspection
### Influence points  
```{r echo=TRUE,size="tiny"}
inflm.HC_LM_1 <-  influence.measures(HC_LM_1)
which(apply(inflm.HC_LM_1$is.inf, 1, any))
# which observations 'are' influential
dim(summary(inflm.HC_LM_1)) # only these
```

### Influence points analysis

* n= 4 and k= 9

* A Point is an influence point if 

    + COOK's D-statistic(D) > 1 
    
    + DFBETAS > ${2 \over \sqrt{n}}  = {2 \over \sqrt{4}}   = 1$, i.e DFBETAS > 1
    
    + DFFITS > $2 * \sqrt{ k \over n} = 2 * \sqrt{9 \over 4} = 3$ , i.e DFFITS > 3
  
  
 * INFLUENCE POINT 
     + Point 3 (as per DFBETAS). None as per DFFITS and cook's distance D.
 
 * INVESTIGATION 
     + Point 3 is within 2 sigma limits of X1,X2,X3 and X4. So retain it
 
 
```{r  include=FALSE}
# Find 3 sigma limits of X1
mean(haldcement$X1) - 3 * sd(haldcement$X1); mean(haldcement$X1) + 3 * sd(haldcement$X1);
# Find 3 sigma limits of X2
mean(haldcement$X2) - 3 * sd(haldcement$X2);mean(haldcement$X2) + 3 * sd(haldcement$X2)
# Find 3 sigma limits for X3
mean(haldcement$X3) - 3 * sd(haldcement$X2);mean(haldcement$X3) + 3 * sd(haldcement$X3)
# Find 3 sigma limits for X4
mean(haldcement$X4) - 3 * sd(haldcement$X4);mean(haldcement$X4) + 3 * sd(haldcement$X4)

# Find 2 sigma limits of X1
mean(haldcement$X1) - 2 * sd(haldcement$X1); mean(haldcement$X1) + 2 * sd(haldcement$X1);
# Find 2 sigma limits of X2
mean(haldcement$X2) - 2 * sd(haldcement$X2);mean(haldcement$X2) + 2 * sd(haldcement$X2)
# Find 2 sigma limits for X3
mean(haldcement$X3) - 2 * sd(haldcement$X2);mean(haldcement$X3) + 2 * sd(haldcement$X3)
# Find 2 sigma limits for X4
mean(haldcement$X4) - 2 * sd(haldcement$X4);mean(haldcement$X4) + 2 * sd(haldcement$X4)
```

## Multicollinearity inspection

**Variance inflation factors (VIF):**

the variance inflation factor for the j th explanatory variable is defined
as $VIF_j = 1\over (R^{2}_j)

One or more large VIFs indicate the presence of multicollinearity in the data.
**In practice, usually a VIF > 5 or 10 indicates that the associated regression coefficients are poorly estimated because of multicollinearity.**

### Multicollinearity Check
```{r echo=FALSE}
vif(HC_LM_1)
```

* OBSERVATION:  SEVERE MULTICOLLINEARITY as VIF > 10 FOR ALL VARIABLES

### Correlation Co-efficients - Pearson test
```{r include=FALSE}
# We can also have pearson correlation coeffcients along with p-values.
library("Hmisc")
haldcement_cormat <- rcorr(as.matrix(haldcement))
# ++++++++++++++++++++++++++++
# flattenCorrMatrix
# ++++++++++++++++++++++++++++
# cormat : matrix of the correlation coefficients
# pmat : matrix of the correlation p-values
flattenCorrMatrix <- function(cormat, pmat) {
  ut <- upper.tri(pmat)  # ut <- upper.tri(cormat) 
  data.frame(
    row = rownames(cormat)[row(cormat)[ut]],
    column = rownames(cormat)[col(cormat)[ut]],
    cor  =(cormat)[ut],
    p = pmat[ut]
  )
}

```
```{r echo=FALSE,size="tiny"}
# Actual Output
flattenCorrMatrix(haldcement_cormat$r, haldcement_cormat$P)
```

* Test of Hypothesis (Pearson)
    + H0: $\rho$ = 0 ; i.e No relationship between the variables.
    + H1 : $\rho$ $\neq$ 0

*  OBSERVATION:
    + No Correllation : Accept H0 : X1,X2 (p=0.4); X2,X3 (p=0.6),X1,X4(p=0.4),X3,X4 (p=0.9) 
    + Correlation exists : Accept H1 : X1,X3 (< 0.05);X2,X4 (< 0.05)

##  Feature Engineering ( Feature Selection)
###  Variable Selection - Forward selection
```{r  include = FALSE}
###################
#Iteration-1
#Forward Selection
FSR=regsubsets(Y~.,data=haldcement,method="forward")
summary(FSR)
Modelsummary=cbind(summary(FSR)$which,R2=summary(FSR)$rsq,SSres=summary(FSR)$rss,AdjR2=summary(FSR)$adjr2,Cp=summary(FSR)$cp,BIC=summary(FSR)$bic)
```
```{r echo=FALSE}
Modelsummary
```

*  RECOMMENDATION :  X4 +X1 =97%

###  Variable Selection - Backward Elimination
```{r  include = FALSE}
#***************************
#Backward elimination
BER=regsubsets(Y~.,data=haldcement,method="backward")
Modelsummary=cbind(summary(BER)$which,R2=summary(BER)$rsq,SSres=summary(BER)$rss,AdjR2=summary(BER)$adjr2,Cp=summary(BER)$cp,BIC=summary(BER)$bic)
```
```{r  echo = FALSE}
Modelsummary
```
*  RECOMMENDATION :  X2 + X1 = 97%


###  Variable Selection - Stepwise Regression
```{r  include = FALSE}
#***************************
#Stepwise Regression
SWR=regsubsets(Y~.,data=haldcement,method="seqrep")
Modelsummary=cbind(summary(SWR)$which,R2=summary(SWR)$rsq,SSres=summary(SWR)$rss,AdjR2=summary(SWR)$adjr2,Cp=summary(SWR)$cp,BIC=summary(SWR)$bic)

```

```{r  echo = FALSE}
Modelsummary
```

*  RECOMMENDATION : X4+X1+X2= 97%

###  Variable Selection - PCA
```{r include=FALSE}
haldcement_x <- haldcement[,c(-1)]

#Scale the data
res.pca <- prcomp(haldcement_x, center = TRUE, scale = TRUE)
summary(res.pca)
library(factoextra)

```
```{r echo=FALSE}
# Contributions of variables to PC1
fviz_contrib(res.pca, choice = "var", axes = 1, top = 10)
fviz_contrib(res.pca, choice = "var", axes = 2, top = 10)
```

*  RECOMMENDATION : : PCA1: X2 and X4 ( X4 very close to X3)

##  Model Selection
###  Finalizing Variable/Model selection 

```{r include=FALSE}
### Model 1 - X1 + X2 - stepwise regression
HC_LM_Model1 = lm(Y~X1+X2)
#### Model 2 - X2 and X4  - PCA 1
HC_LM_X2_X4 <- lm(Y~X2+X4)
#######  Model 3 - X4 + X1
HC_LM_X4_X1 <- lm(Y~ X4+X1 )
######## Model 4 - X2 and X3 ( PCA1 - slightly modified as X2 and X4 are correlated)
HC_LM_X2_X3 <- lm(Y~X2+X3)
```

```{r include=FALSE,size="tiny'"}
PRESS(HC_LM_Model1)
vif(HC_LM_Model1)
summary(HC_LM_Model1)$adj.r.squared
PRESS(HC_LM_X2_X4)
vif(HC_LM_X2_X4)
summary(HC_LM_X2_X4)$adj.r.squared
PRESS(HC_LM_X4_X1)
vif(HC_LM_X4_X1)
summary(HC_LM_X4_X1)$adj.r.squared
PRESS(HC_LM_X2_X3)
vif(HC_LM_X2_X3)
summary(HC_LM_X2_X3)$adj.r.squared
```

VAR  |PRESS |VIF|$R^{2}$(adj)| $R^{2}$(pred)| 
-----|------|---|------------|--------------|
X1+X2|93.8  |1  |   97%      |      96%     |
X2+X4|1461.8|18 |   61%      |      46%     |
X4+X1|121.2 |1  |   96%      |      95%     |
X2+X3|701.7 |1  |   81%      |      74%     |


*  RECOMMENDATION :PRESS stat should be small ( so that $R^{2}(pred) is high) and VIF=1. 
    + We will go with Model 1 : X1+X2   

##  Model Adequacy checking
The fitting of linear regression model, estimation of parameters testing of hypothesis properties of the estimator are based on following major assumptions:
1. The relationship between the study variable and explanatory variables is linear, atleast approximately. (L)
2. The error term has zero mean.
3. The error term has constant variance.(E)
4. The errors are uncorrelated. (I)
5. The errors are normally distributed.(N)

LINE 
* Linearity
* Independence
* Normality
* Equality of Variance

###  FINAL Model : X1 + X2 
```{r include=FALSE}
Model_LM_final = lm(haldcement$Y ~ X1+X2, data = haldcement )
summary(Model_LM_final)
```
####  Linearity Test 
```{r echo=FALSE}
### 1.  L - Linearity test - e vs X1, e vs X2
par(mfrow=c(2,2))
##### e vs X1
plot(haldcement$X1,rstudent(Model_LM_final),xlab = "X1",ylab = "rstudent(ti)")
##### e vs X2
plot(haldcement$X2,rstudent(Model_LM_final),xlab = "X2",ylab = "rstudent(ti)")
######### OBSERVATION : Seems to be in a band.
```

####  Independence Test 
```{r echo=FALSE}
#### 2.  (I) - Independence/AUTO CORRELATION TEST
#The Durbin-Watson (D-W) test is used for testing the hypothesis of lack of first order autocorrelation in the
#disturbance term. The null hypothesis is
# H0 : ρ = 0
#positive autocorrelation of e t ’s ⇒ d < 2
#negative autocorrelation of e t ’s ⇒ d > 2
#zero autocorrelation of e t ’s ⇒ d ≈ 2
durbinWatsonTest(Model_LM_final)
#######RESULT
#lag Autocorrelation D-W Statistic p-value
#1     -0.05450402       1.92164   0.892
#Alternative hypothesis: rho != 0
######## p is high so NULL flies . INDEPENDENCE TEST PASSES. No AUTOCORRELATION issue.
```
####  Normality Test 
```{r echo=FALSE}
####################### 1. N - Normality Significance test for Residuals.
# H0 : Part of Normal distribution
# H1 : Not part of normal distribution
shapiro.test(rstandard(Model_LM_final))
#####Result 
#Shapiro-Wilk normality test
#data:  rstandard(Model_LM_final)
#W = 0.90219, p-value = 0.1433
####### We can also use Anderson-Darling test
#H0: The data follows the normal distribution
#H1: The data do not follow the normal distribution
ad.test(rstandard(Model_LM_final))
####Result
#Anderson-Darling normality test
#data:  rstandard(Model_LM_final)
#A = 0.61361, p-value = 0.08628
#########Normal Probability plot
ors <- rstandard(Model_LM_final)
# normal probability plot is also known as QQ plot in R ? Seems to be based on quantiles and median.. double check
qqnorm(ors, pch = 1, frame = FALSE)
qqline(ors, col = "steelblue", lwd = 2)
#### RESDUALS FOLLOW NORMAL DISTRIBUTION.
```
####  Equality of Variance Test 
```{r echo=FALSE}
####################### 2. (E) -  HOMOSCADASTICITY TEST
### NULL hypothesis H 0 : σ 1 2 = σ 2 2 = ... = σ n 2
bptest(Model_LM_final)
######Results
#studentized Breusch-Pagan test
#data:  Model_LM_final
#BP = 0.16688, df = 2, p-value = 0.9199
#### RESDUALS have EQUAL VARIANCE .. HOMOSCADASTICTY OK.
```