---
title: "Assignment-9"
format: html
editor: visual
---

## Principal Component Analysis (PCA) using Hald Cement Dataset

### Objective

1.  To understand Principal Component Analysis (PCA).

2.  To reduce dimensionality of correlated variables.

3.  To identify principal components explaining maximum variance.

4.  To visualize relationships between variables using PCA plots.

# Dataset

Hald Cement Dataset

Variables used in the dataset:

| Variable | Description                   |
|----------|-------------------------------|
| X1       | Tricalcium aluminate          |
| X2       | Tricalcium silicate           |
| X3       | Tetracalcium aluminoferrite   |
| X4       | Dicalcium silicate            |
| Y        | Heat evolved during hardening |

For PCA we will analyze only the predictor variables **X1, X2, X3, X4**.

```{r}
library(corrplot)
library(factoextra)
library(GGally)
```

### Loading Data Set

```{r}
setwd("D:/Sem-2/R-/7,8,9,10/7")
data<-read.csv("BEST-Hald Cement data.csv")
print(data)
```

```{r}
str(data)
summary(data)
```

```{r}
cor(data)
```

### Select Predictor Variable

PCA Should only be applied to predictor Variables

```{r}
predictors <- data[,c("X1","X2","X3","X4")]
```

### Scatterplot Matrix

```{r}
pairs(predictors,
main="Scatterplot Matrix of Cement Variables")
```

### Correlation Matrix

```{r}
cor_matrix <- cor(predictors)

cor_matrix
```

### Correlation Heatmap

```{r}
corrplot(cor_matrix,
method="color",
type="upper",
tl.col="black",
tl.srt=45)
```

### Standardize Varialbles

-   PCA requires Variables to be standardized

```{r}
scaled_data <- scale(predictors)
```

### Doing PCA now

```{r}
pca_model <- prcomp(scaled_data)

summary(pca_model)
```

-   Eigen values here represents the variance explained by each principle component

```{r}
pca_model$sdev^2
```

### shows how each variable contributes to component

-   PCA Loading

    ```{r}
    pca_model$rotation
    ```

### Screen Plot

```{r}
plot(pca_model,
type="l",
main="Scree Plot")
```

### Eigenvalue Visualization

```{r}
fviz_eig(pca_model,
addlabels=TRUE)
```

### Variable Contribution PLot

```{r}
fviz_pca_var(pca_model,
col.var="contrib")
```

```{r}
summary(pca_model)$importance
```

### Observation Plot

```{r}
fviz_pca_ind(pca_model)
```

### Extracting Principal Components

```{r}
pc_scores <- pca_model$x

head(pc_scores)
```

### Visualizing first two components

```{r}
plot(pc_scores[,1],
pc_scores[,2],
xlab="PC1",
ylab="PC2",
main="Principal Component Space")
```

```{r}
var_explained <- pca_model$sdev^2 / sum(pca_model$sdev^2)

plot(cumsum(var_explained),
     type="b",
     xlab="Principal Components",
     ylab="Cumulative Variance Explained",
     main="Cumulative Variance Plot")
```

## Relation ship betwwn PC1 and target variable

```{r}
plot(pc_scores[,1], data$Y,
     xlab="PC1",
     ylab="Heat (Y)",
     main="PC1 vs Heat Evolution")
```
