---
title: "Assignmnet-10"
author: "Tushar Sharma"
format: pdf
editor: visual
---

### Load Libraries

```{r}
library(mlbench) # PimaIndiansDiabetes dataset
library(caret) # createDataPartition(), confusionMatrix()
library(pROC) # roc(), auc(), coords()
library(ggplot2)
library(dplyr)


```

### Load Dataset

```{r}
data(PimaIndiansDiabetes)
df <- PimaIndiansDiabetes
head(df)

```

```{r}
str(df)
summary(df)
```

### Data Exploration

#### Class Distribution

```{r}
table(df$diabetes)
cat(sprintf("\nClass balance: %.1f%% negative | %.1f%% positive\n",
500/768*100, 268/768*100))

```

### Glucose Distribution by Diabetes Class

```{r}
ggplot(df, aes(x = glucose, fill = diabetes)) +
geom_histogram(alpha = 0.6, bins = 30, colour = "white") +
scale_fill_manual(values = c("neg" = "#F48FB1", "pos" = "#4DD0E1")) +
theme_minimal(base_size = 12) +
labs(title = "Glucose Distribution by Diabetes Class",
x = "Glucose Level",
y = "Count",
fill = "Diabetes")
```

### Additonal Exploratory Plots

```{r}
ggplot(df, aes(x = diabetes, y = mass, fill = diabetes)) +
geom_boxplot(alpha = 0.7, outlier.colour = "red", outlier.size = 1.5) +
scale_fill_manual(values = c("neg" = "#81D4FA", "pos" = "#EF9A9A")) +
theme_minimal(base_size = 12) +
labs(title = "BMI by Diabetes Class",
x = "Diabetes Status",
y = "BMI (mass)",
fill = "Diabetes")
```

```{r}
ggplot(df, aes(x = diabetes, y = age, fill = diabetes)) +
geom_boxplot(alpha = 0.7, outlier.colour = "red", outlier.size = 1.5) +
scale_fill_manual(values = c("neg" = "#C5E1A5", "pos" = "#FFCC80")) +
theme_minimal(base_size = 12) +
labs(title = "Age Distribution by Diabetes Class",
x = "Diabetes Status",
y = "Age",
fill = "Diabetes")
```

```{r}
train_index <- createDataPartition(df$diabetes,
p = 0.8,
list = FALSE)
train <- df[train_index, ]
test <- df[-train_index, ]
cat(sprintf("Training rows : %d\n", nrow(train)))

```

```{r}
cat(sprintf("Testing rows : %d\n", nrow(test)))
cat(sprintf("\nTrain class distribution:\n"))
print(table(train$diabetes))
```

```{r}
model <- glm(diabetes ~ .,
data = train,
family = binomial)
summary(model)
```

### Generate Predicted Probabilities

```{r}
prob_predictions <- predict(model,
newdata = test,
type = "response")
head(prob_predictions)
cat(sprintf("\nProbability range: [%.4f, %.4f]\n",
min(prob_predictions), max(prob_predictions)))
```

## ROC Curve Object

```{r}
roc_obj <- roc(test$diabetes,
              prob_predictions,
              levels = c("neg", "pos"))
print(roc_obj)
```

### AUC Calculation

```{r}
auc_value <- auc(roc_obj)
cat(sprintf("AUC: %.4f\n", auc_value))
# AUC confidence interval
auc_ci <- ci.auc(roc_obj)
cat(sprintf("95%% CI: [%.4f, %.4f]\n", auc_ci[1], auc_ci[3]))
```

### ROC Curve Visualisation

```{r}
plot(roc_obj,
col = "blue",
lwd = 3,
main = "ROC Curve for Diabetes Prediction")
abline(a = 0, b = 1, col = "gray", lty = 2)
legend("bottomright",
legend = sprintf("AUC = %.4f", auc_value),
col = "blue",
lwd = 3,
bty = "n")
```

### 

### ggplot2 ROC Curve

```{r}
roc_df <- data.frame(
FPR = 1 - roc_obj$specificities,
TPR = roc_obj$sensitivities,
Thresholds = roc_obj$thresholds
)
ggplot(roc_df, aes(x = FPR, y = TPR)) +
geom_line(colour = "steelblue", linewidth = 1.2) +
geom_abline(slope = 1, intercept = 0,
colour = "gray50", linetype = "dashed") +
geom_area(fill = "steelblue", alpha = 0.1) +
annotate("text", x = 0.65, y = 0.15,
label = sprintf("AUC = %.4f", auc_value),
size = 5, colour = "steelblue", fontface = "bold") +
scale_x_continuous(limits = c(0, 1), labels = scales::percent_format()) +
scale_y_continuous(limits = c(0, 1), labels = scales::percent_format()) +
theme_minimal(base_size = 12) +
labs(title = "ROC Curve — Diabetes Prediction (Logistic Regression)",
## ── Optimal Threshold ──
## threshold sensitivity specificity
## 1 0.2828958 0.8867925 0.77
subtitle = "Pima Indians Diabetes Dataset",
x = "False Positive Rate (1 − Specificity)",
y = "True Positive Rate (Sensitivity)")

```

### Optimal Threshold

```{r}
best_threshold <- coords(roc_obj,
"best",
ret = c("threshold", "sensitivity", "specificity"))
print(best_threshold)
```

### Probability Distribution

```{r}
ggplot(data.frame(prob = prob_predictions,
actual = test$diabetes),
aes(x = prob, fill = actual)) +
geom_histogram(bins = 30, colour = "white", alpha = 0.75,
position = "identity") +
geom_vline(xintercept = 0.5, colour = "red",
linetype = "dashed", linewidth = 1) +
geom_vline(xintercept = best_threshold$threshold,
colour = "darkgreen", linetype = "dashed", linewidth = 1) +
scale_fill_manual(values = c("neg" = "#81D4FA", "pos" = "#EF9A9A")) +
theme_minimal(base_size = 12) +
labs(title = "Distribution of Predicted Probabilities",
subtitle = "Red dashed = 0.5 threshold | Green dashed = optimal threshold",
x = "Predicted Probability",
y = "Frequency",
fill = "Actual Class")
```

### Model Evalutation

### Confusion Matrix at degault Threshhold

```{r}
predicted_class_05 <- ifelse(prob_predictions > 0.5, "pos", "neg")
cm_05 <- confusionMatrix(as.factor(predicted_class_05), test$diabetes)
print(cm_05)
```

### Confusion Matrix at Optimal Threshold

```{r}
predicted_class_opt <- ifelse(prob_predictions > best_threshold$threshold,
"pos", "neg")
cm_opt <- confusionMatrix(as.factor(predicted_class_opt), test$diabetes)
print(cm_opt)

```

### 
