---
title: "Regression Analysis Case Study - Hald Cement Dataset"
author: "G Radhesyam"
date: "19 March 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```
```{r include=FALSE}
# Set the environment
getwd()
# Please set your current working directory path here
setwd("E:/Radhesyam/RProjects/mdsc206/Regression")

#Load the libraries required.
library(tidyverse)
library(MASS)
library(ggplot2) # plot
library(calibrate) # textxy??
library(car)#vif, durbin-watson test
library(orcutt)  # To fix atuto-correlation issue
library(leaps) #regsubset - forward / Backward Substitution
library(qpcR)#To calculate PRESS Residual and PRESS Statistic
library(factoextra)## PCA graphs
library(nortest) #### for normality testing - Anderson-Darling etc
library(Hmisc)
```

## HALD CEMENT Dataset

* The response variable Y is the heat evolved in a cement mix.

* The four explanatory variables are ingredients of the mix,
  + X1:Tricalcium aluminate
  + X2:Tricalcium silicate,
  + X3:Tetracalcium alumino ferrite
  + X4:Dicalcium silicate.

## HALD CEMENT Dataset
```{r echo=FALSE}
haldcement <- readr::read_csv("BEST-Hald Cement data.csv")
head(haldcement ) # fetch first 6 rows
str(haldcement ) #
spec(haldcement) # spec() extracts the full column specification from a tibble created by readr.
attach(haldcement) # So that we can directly refer to column names/feature names
```

## HALD CEMENT - FINAL MODEL SUMMARY
* Model : X1+X2
  - Y = 52.5773  + (1.4683) X1 + ( 0.6623) X2
  - All Regression assumptions met.

* Normality test
  + Shapiro-wilks Test - Distribution is Normal
 - W = 0.90219, p-value = 0.1433
  + Anderson-Darling test - Distribution is normal
 - A = 0.61361, p-value = 0.08628


* Equal Variance Test
  + studentized Breusch-Pagan test
  - BP = 0.16688, df = 2, p-value = 0.9199
  - H0 : Variances are equal

## HALD CEMENT - FINAL MODEL SUMMARY (Contd...)
      * Independence/Auto Correllation Test (Durbin-Watson (D-W) test )
         - H0 : Auto correlation is zero

lag  | Autocorrelation | D-W Statistic | p-value
-----|-----------------|---------------|--------
1   | -0.05450402     |  1.92164      | 0.892


* Linearity test
  + Box-Tidwell - H0 : No Transformation needed
  + Box-Tidwell test (Y,X1)
  +  MLE of lambda Score Statistic (z) Pr(>|z|)
  +       0.23145             -0.7366   0.4614
  + Box-Tidwell test ( Y,X2)
  + MLE of lambda Score Statistic (z) Pr(>|z|)
  +    -0.71695               -1.35    0.177

## HALD CEMENT - MODEL ANALYSIS and DETAILS
* An F statistic is a value you get when you run an ANOVA test or a regression analysis to find out if the means
  between two populations are significantly different.
* ItŌĆÖs similar to a T statistic from a T-Test;
* A T-test will tell you if a single variable is statistically significant and
* an F test will tell you if a group of variables are jointly significant.
* https://www.statisticshowto.com/probability-and-statistics/f-statistic-value-test/
* Regression NULL Hypothesis:  All of the regression coefficients are equal to zero
* Tip: P-value is High ,NUll will Fly
*      P-value is low, NULL will blow
* T-Test in Regression
* https://www.statology.org/t-test-linear-regression/
```{r echo=FALSE}
# So let us build an initial model with Y vs X1,X2,X3 and X4
HC_LM_1 <- lm(haldcement$Y ~ ., data = haldcement )

# What's the model - linear equation
#HC_LM_1

#Summary of the model ( R2 adjusted etc)
summary(HC_LM_1)
```
## HALD CEMENT - Scatter plot
```{r echo=FALSE}
library(ggplot2)
panel.cor <- function(x, y){
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- round(cor(x, y), digits=2)
  txt <- paste0("R = ", r)
  cex.cor <- 0.8/strwidth(txt)
  text(0.5, 0.5, txt, cex = cex.cor * r)
}
pairs(haldcement$Y ~ ., data = haldcement,lower.panel = panel.cor)
```

* https://ggobi.github.io/ggally/reference/ggpairs.html
*
```{r echo=FALSE}
# Quick display of two cabapilities of GGally, to assess the distribution and correlation of variables
library(GGally)
ggpairs(data=haldcement, title="Haldcement Dataset correlogram with ggpairs()",
        upper = list(continuous = wrap("cor", size = 9)))
```
```{r echo=FALSE}
# Quick display of two cabapilities of GGally, to assess the distribution and correlation of variables
library(GGally)
# Nice visualization of correlations
ggcorr(data=haldcement, method = c("everything", "pearson"))
```

## HALD CEMENT - Influence points
* https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/influence.measures
* influence.measures: Regression Deletion Diagnostics
* https://rpubs.com/mpfoley73/501093
```{r echo=FALSE,size="tiny"}
inflm.HC_LM_1 <-  influence.measures(HC_LM_1)
which(apply(inflm.HC_LM_1$is.inf, 1, any))
# which observations 'are' influential
dim(summary(inflm.HC_LM_1)) # only these
```

## Influence points analysis

* n= 4 and k= 9

* A Point is an influence point if

  + COOK's D-statistic(D) > 1

  + DFBETAS > ${2 \over \sqrt{n}}  = {2 \over \sqrt{4}}   = 1$, i.e DFBETAS > 1

  + DFFITS > $2 * \sqrt{ k \over n} = 2 * \sqrt{9 \over 4} = 3$ , i.e DFFITS > 3


 * INFLUENCE POINT
  + Point 3 (as per DFBETAS). None as per DFFITS and cook's distance D.

 * INVESTIGATION
  + Point 3 is within 2 sigma limits of X1,X2,X3 and X4. So retain it


```{r  include=FALSE}
# Find 3 sigma limits of X1
paste0("3 sigma limits of X1  ",(mean(haldcement$X1) - 3 * sd(haldcement$X1)),"  ",(mean(haldcement$X1) + 3 * sd(haldcement$X1)))
# Find 3 sigma limits of X2
paste0("3 sigma limits of X2  ",(mean(haldcement$X2) - 3 * sd(haldcement$X2)),"  ",(mean(haldcement$X2) + 3 * sd(haldcement$X2)))
# Find 3 sigma limits for X3
paste0("3 sigma limits for X3  ",(mean(haldcement$X3) - 3 * sd(haldcement$X2)),"  ",(mean(haldcement$X3) + 3 * sd(haldcement$X3)))
# Find 3 sigma limits for X4
paste0("3 sigma limits for X4  ",(mean(haldcement$X4) - 3 * sd(haldcement$X4)),"  ",(mean(haldcement$X4) + 3 * sd(haldcement$X4)))

# Find
paste0("2 sigma limits of X1  ",(mean(haldcement$X1) - 2 * sd(haldcement$X1))," ",(mean(haldcement$X1) + 2 * sd(haldcement$X1)))
# Find 2 sigma limits of X2
paste0("2 sigma limits of X2  ",(mean(haldcement$X2) - 2 * sd(haldcement$X2)),"  ",(mean(haldcement$X2) + 2 * sd(haldcement$X2)))
# Find 2 sigma limits for X3
paste0("2 sigma limits for X3  ",(mean(haldcement$X3) - 2 * sd(haldcement$X3)),"  ",(mean(haldcement$X3) + 2 * sd(haldcement$X3)))
# Find 2 sigma limits for X4
paste0("2 sigma limits for X4  ",(mean(haldcement$X4) - 2 * sd(haldcement$X4)),"  ",(mean(haldcement$X4) + 2 * sd(haldcement$X4)))
```

## MULTICOLLINEARITY Check
```{r echo=FALSE}
car::vif(HC_LM_1)
```

* OBSERVATION:  SEVERE MULTICOLLINEARITY as VIF > 10 FOR ALL VARIABLES


## Correlation Co-efficients - Pearson test
```{r include=FALSE}
############################## REMEDY FOR MULTICOLLINEARITY ######################
# We can also have pearson correlation coeffcients along with p-values.
library("Hmisc")
haldcement_cormat <- rcorr(as.matrix(haldcement))
# ++++++++++++++++++++++++++++
# flattenCorrMatrix
# ++++++++++++++++++++++++++++
# cormat : matrix of the correlation coefficients
# pmat : matrix of the correlation p-values
flattenCorrMatrix <- function(cormat, pmat) {
  ut <- upper.tri(pmat)  # ut <- upper.tri(cormat)
  data.frame(
    row = rownames(cormat)[row(cormat)[ut]],
    column = rownames(cormat)[col(cormat)[ut]],
    cor  =(cormat)[ut],
    p = pmat[ut]
  )
}

```
```{r echo=FALSE,size="tiny"}
# Actual Output
flattenCorrMatrix(haldcement_cormat$r, haldcement_cormat$P)
```

* Test of Hypothesis (Pearson)
  + H0: $\rho$ = 0 ; i.e No relationship between the variables.
  + H1 : $\rho$ $\neq$ 0

*  OBSERVATION:
 + No Correllation : Accept H0 : X1,X2 (p=0.4); X2,X3 (p=0.6),X1,X4(p=0.4),X3,X4 (p=0.9)
 + Correlation exists : Accept H1 : X1,X3 (< 0.05);X2,X4 (< 0.05)


##  Variable Selection - Forward selection
```{r  include = FALSE}
###################
#Iteration-1
#Forward Selection
FSR=regsubsets(Y~.,data=haldcement,method="forward")
summary(FSR)
Modelsummary=cbind(summary(FSR)$which,R2=summary(FSR)$rsq,SSres=summary(FSR)$rss,AdjR2=summary(FSR)$adjr2,Cp=summary(FSR)$cp,BIC=summary(FSR)$bic)
```
```{r echo=FALSE}
Modelsummary
```

*  RECOMMENDATION :  X4 +X1 =97%

##  Variable Selection - Backward Elimination
```{r  include = FALSE}
#***************************
#Backward elimination
BER=regsubsets(Y~.,data=haldcement,method="backward")
Modelsummary=cbind(summary(BER)$which,R2=summary(BER)$rsq,SSres=summary(BER)$rss,AdjR2=summary(BER)$adjr2,Cp=summary(BER)$cp,BIC=summary(BER)$bic)
```
```{r  echo = FALSE}
Modelsummary
```


*  RECOMMENDATION :  X2 + X1 = 97%


##  Variable Selection - Stepwise Regression
```{r  include = FALSE}
#***************************
#Stepwise Regression
SWR=regsubsets(Y~.,data=haldcement,method="seqrep")
Modelsummary=cbind(summary(SWR)$which,R2=summary(SWR)$rsq,SSres=summary(SWR)$rss,AdjR2=summary(SWR)$adjr2,Cp=summary(SWR)$cp,BIC=summary(SWR)$bic)

```

```{r  echo = FALSE}
Modelsummary
```

*  RECOMMENDATION : X4+X1+X2= 97%

##  Variable Selection - PCA
```{r include=FALSE}
haldcement_x <- haldcement[,c(-1)]

#Scale the data
res.pca <- prcomp(haldcement_x, center = FALSE, scale = TRUE)
summary(res.pca)
library(factoextra)


```
```{r echo=FALSE}
# Contributions of variables to PC1
fviz_contrib(res.pca, choice = "var", axes = 1, top = 10)
```
*  RECOMMENDATION : : PCA1: X2 and X4 ( X4 very close to X3)


##  FINALIZING VARIABLE/MODEL selection

```{r include=FALSE}
### Model 1 - X1 + X2 - stepwise regression
HC_LM_Model1 = lm(Y~X1+X2)
#### Model 2 - X2 and X4  - PCA 1
HC_LM_X2_X4 <- lm(Y~X2+X4)
#######  Model 3 - X4 + X1
HC_LM_X4_X1 <- lm(Y~ X4+X1 )
######## Model 4 - X2 and X3 ( PCA1 - slightly modified as X2 and X4 are correlated)
HC_LM_X2_X3 <- lm(Y~X2+X3)
```

```{r include=FALSE,size="tiny'"}
PRESS(HC_LM_Model1)
vif(HC_LM_Model1)
summary(HC_LM_Model1)$adj.r.squared
PRESS(HC_LM_X2_X4)
vif(HC_LM_X2_X4)
summary(HC_LM_X2_X4)$adj.r.squared
PRESS(HC_LM_X4_X1)
vif(HC_LM_X4_X1)
summary(HC_LM_X4_X1)$adj.r.squared
PRESS(HC_LM_X2_X3)
vif(HC_LM_X2_X3)
summary(HC_LM_X2_X3)$adj.r.squared
```

VAR  |PRESS |VIF|$R^{2}$(adj)| $R^{2}$(pred)|
-----|------|---|------------|--------------|
X1+X2|93.8  |1  |   97%      |      96%     |
X2+X4|1461.8|18 |   61%      |      46%     |
X4+X1|121.2 |1  |   96%      |      95%     |
X2+X3|701.7 |1  |   81%      |      74%     |


*  RECOMMENDATION :PRESS stat should be small ( so that R-sq(pred) is high) and VIF=1.
 + We will go with Model 1 : X1+X2

##  FINAL Model : X1 + X2 : Validation of Regression Assumptions
```{r include=FALSE}
################################ FINAL VARIABLE/MODEL selection ##########
############## No clear Winner, as different model gave different combinations. Use PRESS() to break the tie.
### Model 1 - X1 + X2 - stepwise regression
HC_LM_Model1 = lm(Y~X1+X2)
PRESS(HC_LM_Model1)
vif(HC_LM_Model1)
summary(HC_LM_Model1)$adj.r.squared
#### Model 2 - X2 and X4  - PCA 1
HC_LM_X2_X4 <- lm(Y~X2+X4)
PRESS(HC_LM_X2_X4)
vif(HC_LM_X2_X4)
summary(HC_LM_X2_X4)$adj.r.squared
#######  Model 3 - X4 + X1
HC_LM_X4_X1 <- lm(Y~ X4+X1 )
PRESS(HC_LM_X4_X1)
vif(HC_LM_X4_X1)
summary(HC_LM_X4_X1)$adj.r.squared
######## Model 4 - X2 and X3 ( PCA1 - slightly modified as X2 and X4 are correlated)
HC_LM_X2_X3 <- lm(Y~X2+X3)
PRESS(HC_LM_X2_X3)
vif(HC_LM_X2_X3)
summary(HC_LM_X2_X3)$adj.r.squared

######### CONCLUSION
####### Model1 - X1+X2 : PRESS stat : 93.88255, VIF =1, Radj - 97% , R-squared (predicted) - 0.9654305
####### Model2 -X2+X4- PCA 1 : PRESS stat : 1461.814, VIF > 18, Radj - 61%, R-squared (predicted) - 0.4617298
###### Model 3 - X4+X1  : PRESS stat : 121.2244, VIF =1 , Radj - 96% , R-squared (predicted) - 0.9553627
###### Model 4 - X2+X3 : PRESS stat :701.7432 , VIF =1, Radj - 81%, R-squared (predicted) - 0.7416037
#### Since PRESS stat should be small ( so that R-sq(pred) is high), VIF=1. We will go with Model 1 : X1+X2
################################# FINAL MODEL #################
Model_LM_final = lm(haldcement$Y ~ X1+X2, data = haldcement )
summary(Model_LM_final)
####################### 1. N - Normality Significance test for Residuals.
# H0 : Part of Normal distribution
# H1 : Not part of normal distribution
shapiro.test(rstandard(Model_LM_final))
#####Result
#Shapiro-Wilk normality test
#data:  rstandard(Model_LM_final)
#W = 0.90219, p-value = 0.1433
####### We can also use Anderson-Darling test
#H0: The data follows the normal distribution
#H1: The data do not follow the normal distribution
ad.test(rstandard(Model_LM_final))
####Result
#Anderson-Darling normality test
#data:  rstandard(Model_LM_final)
#A = 0.61361, p-value = 0.08628
#########Normal Probability plot
ors <- rstandard(Model_LM_final)
# normal probability plot is also known as QQ plot in R ? Seems to be based on quantiles and median.. double check
qqnorm(ors, pch = 1, frame = FALSE)
qqline(ors, col = "steelblue", lwd = 2)
#### RESDUALS FOLLOW NORMAL DISTRIBUTION.
####################### 2. (E) -  HOMOSCADASTICITY TEST
### NULL hypothesis H 0 : Žā 1 2 = Žā 2 2 = ... = Žā n 2
bptest(Model_LM_final)
######Results
#studentized Breusch-Pagan test
#data:  Model_LM_final
#BP = 0.16688, df = 2, p-value = 0.9199
#### RESDUALS have EQUAL VARIANCE .. HOMOSCADASTICTY OK.
####################### 3.  L - Linearity test - e vs X1, e vs X2
par(mfrow=c(2,2))
##### e vs X1
plot(haldcement$X1,rstudent(Model_LM_final),xlab = "X1",ylab = "rstudent(ti)")
##### e vs X2
plot(haldcement$X2,rstudent(Model_LM_final),xlab = "X2",ylab = "rstudent(ti)")
######### OBSERVATION : Seems to be in a band.
####################### 4.  (I) - Independence/AUTO CORRELATION TEST
#The Durbin-Watson (D-W) test is used for testing the hypothesis of lack of first order autocorrelation in the
#disturbance term. The null hypothesis is
# H0 : Žü = 0
#positive autocorrelation of e t ŌĆÖs ŌćÆ d < 2
#negative autocorrelation of e t ŌĆÖs ŌćÆ d > 2
#zero autocorrelation of e t ŌĆÖs ŌćÆ d Ōēł 2
durbinWatsonTest(Model_LM_final)
#######RESULT
#lag Autocorrelation D-W Statistic p-value
#1     -0.05450402       1.92164   0.892
#Alternative hypothesis: rho != 0
######## p is high so NULL flies . INDEPENDENCE TEST PASSES. No AUTOCORRELATION issue.
####################### FINAL MODEL  ###############
####### Print the current Model
Model_LM_final
#Call:
#  lm(formula = haldcement$Y ~ X1 + X2, data = haldcement)
#Coefficients:
#  (Intercept)           X1           X2
#       52.5773       1.4683       0.6623
# Y = 52.5773  + (1.4683) X1 + ( 0.6623) X2
```


* Normality test
  + Shapiro-wilks Test - Distribution is Normal
 - W = 0.90219, p-value = 0.1433
  + Anderson-Darling test - Distribution is normal
 - A = 0.61361, p-value = 0.08628


* Equal Variance Test
  + studentized Breusch-Pagan test
  - BP = 0.16688, df = 2, p-value = 0.9199

##  FINAL Model : X1 + X2 : Validation of Regression Assumptions (contd..)
* Independence/Auto Correllation Test (Durbin-Watson (D-W) test )

lag  | Autocorrelation | D-W Statistic | p-value
-----|-----------------|---------------|--------
1   | -0.05450402     |  1.92164      | 0.892


* Linearity test
  + Box-Tidwell test (Y,X1)
  +  MLE of lambda Score Statistic (z) Pr(>|z|)
  +       0.23145             -0.7366   0.4614
  + Box-Tidwell test ( Y,X2)
  + MLE of lambda Score Statistic (z) Pr(>|z|)
  +   -0.71695               -1.35    0.177