---
title: "Assignment-2"
author: "Tushar Sharma"
format: pdf
editor: visual
---

**Conditional statements**

1.  if Statement

```{r}
x <- 10
if(x > 5){
print("x is greater than 5")
}
```

```{r}
n <- 7
if(n %% 2 == 1){
print("Number is odd")
}
```

```{r}
marks <- 80
if(marks >= 35){
print("Student passed")
}
```

```{r}
age <- 20
if(age >= 18){
print("Eligible to vote")
}
```

```{r}
temp <- 30
if(temp > 25){
print("Weather is hot")
}
```

2.  if--else Statement

```{r}
x <- 4
if(x %% 2 == 0){
print("Even number")
} else {
print("Odd number")
}
```

```{r}
marks <- 30
if(marks >= 35){
print("Pass")
} else {
print("Fail")
}
```

```{r}
age <- 16
if(age >= 18){
print("Adult")
} else {
print("Minor")
}
```

```{r}
num <- -5
if(num > 0){
print("Positive number")
} else {
print("Negative number")
}
```

```{r}
balance = 15000
if(balance >= 1000){
print("Sufficient balance")
} else {
print("Low balance")
}
```

3.  if--else if--else Statement

```{r}
marks <- 85
if(marks >= 90){
print("Grade A")
} else if(marks >= 75){
print("Grade B")
} else {
print("Grade C")
}
```

```{r}
num <- 0
if(num > 0){
print("Positive")
} else if(num < 0){
print("Negative")
} else {
print("Zero")
}
```

```{r}
day <- "Sunday"
if(day == "Monday"){
print("Working day")
} else if(day == "Sunday"){
print("Holiday")
} else {
print("Normal day")
}
```

```{r}
temp <- 15
if(temp > 30){
print("Hot")
} else if(temp >= 20){
print("Warm")
} else {
print("Cold")
}
```

```{r}
speed <- 90
if(speed > 100){
print("Overspeeding")
} else if(speed >= 60){
print("Normal speed")
} else {
print("Slow speed")
}
```

**LOOPING STATEMENTS**

1.  while Loop

```{r}
i <- 1
while(i <= 5){
print(i)
i <- i + 1
}
```

```{r}
i <- 2
while(i <= 10){
print(i)
i <- i + 2
}
```

```{r}
i <- 5
while(i >= 1){
print(i)
i <- i - 1
}
```

```{r}
count <- 1
while(count <= 3){
print("Hello")
count <- count + 1
}
```

2.  repeat Loop

```{r}
repeat{
print("Executed once")
break
}

```

```{r}
n <- 1
repeat{
print(n^3)
n <- n + 1
if(n > 4) break
}
```

```{r}
x <- 10
repeat{
print(x)
x <- x - 2
if(x <= 0) break
}
```

```{r}
count <- 1
repeat{
print("Loop running")
count <- count + 1
if(count == 4) break
}
```

```{r}
i <- 1
repeat{
print(i)
i <- i + 1
if(i > 5) break
}
```

6.  for Loop

```{r}
for(i in 1:5){
  square <- i * i
  print(paste("Square of", i, "is", square))
}
```

```{r}
for(i in 1:6){
  if(i %% 2 == 0){
    print(paste(i, "is Even"))
  } else {
    print(paste(i, "is Odd"))
  }
}
```

**Functions in R**

1.  Built-in Functions

```{r}

print(sum(1:10))

print(max(c(5, 12, 9, 3)))

print(min(c(15, 7, 20, 4)))

print(mean(10:20))

print(length(c(2, 4, 6, 8, 10)))

```

2.  User defined Function with Argument

```{r}
#Square of a number
square <- function(x){
  return(x * x)
}
print(square(5))
```

```{r}
# Addition of two numbers
add <- function(a, b){
  return(a + b)
}
print(add(10, 20))
```

```{r}
#Area of rectangle
areaRectangle <- function(l, b){
  return(l * b)
}
print(areaRectangle(5, 4))
```

```{r}
#Cube of a number
cube <- function(n){
  return(n^3)
}
print(cube(3))
```

```{r}
#Check even or odd
evenOdd <- function(n){
  if(n %% 2 == 0){
    return("Even")
  } else {
    return("Odd")
  }
}
print(evenOdd(7))
```

3.  Function without argument

```{r}
display <- function(){
  print("Welcome to R Programming")
}
display()
```

```{r}
numbers <- function(){
  for(i in 1:5){
    print(i)
  }
}
numbers()
```

```{r}
showSquare <- function(){
  for(i in 1:4){
    print(i^2)
  }
}
showSquare()
```

```{r}
message <- function(){
  print("This is a user defined function")
}
message()
```

```{r}
simple <- function(){
  print("Function executed successfully")
}
simple()
```

4.  Funcitons with default argument

```{r}
multiply <- function(a = 5, b = 2){
  print(a * b)
}
multiply()
```

```{r}
multiply(4, 3)
```

```{r}
power <- function(x = 2){
  print(x^2)
}
power()
```

```{r}
power(5)
```

5.  Lazy evaluation of function

```{r}
lazy <- function(a, b){
  print(a^2)
}
lazy(4)
```

**Strings\
**

```{r}
# String creation
a <- "Hello"
b <- 'R Programming'
c <- "Learning Strings"
d <- "I'm learning R"
e <- 'Data Science'

print(a)
print(b)
print(c)
print(d)
print(e)

# paste() function
print(paste("Hello", "World"))
print(paste("R", "Programming", sep = "-"))
print(paste("Data", "Science", sep = ""))
print(paste("Score is", 95))
print(paste("Welcome", "to", "R"))

# String functions
print(nchar("Programming"))
print(toupper("r language"))
print(tolower("DATA SCIENCE"))
print(substring("Programming", 1, 4))
print(format(123.4567, nsmall = 2))

```

Date: 21-01-2026

## Charts and Graphs in R

```{r}
#pie chart
slices <- c(10,20,30,40)
labels <- c("A","B","C","D")
pie(slices,labels=labels,main = "1st example")
```

#### Another pie chart with percentage

```{r}
pct <- round(slices/sum(slices)*100)
labels_pct <- paste(lables,pct,"%")
pie(slices,lables=labels_pct,main="2nd with ex. with %")
```

```{r}
marks <- c(10, 20, 30, 40)
labels <- c("Fail", "Pass", "Second", "First")

pie(marks,
    labels = labels,
    main = "Result Distribution")
```

```{r}
students <- c(50, 30, 20)
labels <- c("Hostel", "", "PG")

pie(students,
    labels = labels,
    col = c("skyblue", "lightgreen", "pink"),
    radius = 1,
    main = "Student Residence Type")


```

```{r}
hours <- c(8, 5, 3, 4)
labels <- c("Sleep", "Study", "College", "Sports")

colors <- c("skyblue", "lightgreen", "orange", "pink", "yellow")

pie(hours,
    labels = labels,
    col = colors,
    main = "Daily Time Distribution")

```

```{r}


install.packages("plotrix")  
library(plotrix)

hours <- c(8, 5, 3, 4)
labels <- c("Sleep", "Study", "College", "Sports")

colors <- c("skyblue", "lightgreen", "orange", "pink")

pie3D(hours,
      labels = labels,
      col = colors,
      explode = c(0, 0.25, 0, 0),   # Study highlighted
      main = "Daily Time Distribution (3D Pie Chart)")

```

#### Bar Graphs

```{r}
subjects <- c("Statistics", "ML", "NLP", "Python", "R")
marks <- c(18, 22, 20, 25, 15)

barplot(marks,
        names.arg = subjects,
        col = "skyblue",
        main = "Internal Marks by Subject",
        ylab = "Marks (out of 30)",
        xlab = "Subjects")

text(x = seq_along(marks),
     y = marks,
     labels = marks,
     pos = 3)
```

```{r}
activities <- c("Study", "Sleep", "Sports")
hours <- c(6, 8, 2)

barplot(hours,
        names.arg = activities,
        col = c("orange", "skyblue", "pink"),
        main = "Daily Activities",
        ylab = "Hours")

```

```{r}


days <- c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")
hours <- c(3, 4, 2, 5, 4, 6, 2)

max_index <- which.max(hours)

colors <- rep("lightgray", length(hours))
colors[max_index] <- "darkgreen"

barplot(hours,
        names.arg = days,
        col = colors,
        main = "Study Hours During the Week",
        ylab = "Hours",
        xlab = "Days")

text(x = seq_along(hours),
     y = hours,
     labels = hours,
     pos = 3)

```

```{r}
subjects <- c("Statistics", "ML", "NLP", "Python", "R", "AI")
marks <- c(68, 72, 70, 75, 65, 73)

bp <- barplot(marks,
              names.arg = subjects,
              col = "lightblue",
              ylim = c(0, 80),
              main = "Marks in Different Subjects",
              ylab = "Marks")

text(x = bp,
     y = marks,
     labels = marks,
     pos = 3)

grid(nx = NA, ny = NULL)


```

```{r}
activities <- c("play ", "study")

weekday <- c(3, 1)
weekend <- c(5, 3)

hours <- rbind(weekday, weekend)

barplot(hours,
        beside = TRUE,
        col = c("lightgreen", "pink"),
        names.arg = activities,
        main = "Weekday vs Weekend")

legend("topleft",
       legend = c("Weekday", "Weekend"),
       fill = c("lightgreen", "pink"))

```

#USing DATA FRAMES

```{r}
# data frame
marks_df <- data.frame(
  Subject = c("Statistics", "ML", "NLP", "Python", "R", "AI"),
  Marks   = c(68, 72, 70, 75, 65, 73)
)

View(marks_df)

```

```{r}
barplot(marks_df$Marks,
        names.arg = marks_df$Subject,
        ylim = c(0, 100),
              main = "Marks in Different Subjects",
              ylab = "Marks")

text(x = bp,
     y = marks,
     labels = marks,
     pos = 3)

grid(nx = NA, ny = NULL)

```

#### BoxPlots

```{r}
stats <- c(65, 70, 72, 68, 75)
ml    <- c(60, 62, 65, 70, 68)
python<- c(55, 58, 60, 63, 65)

data <- list(Statistics = stats,
             ML = ml,
             Python = python)

boxplot(data,
        col = c("skyblue", "lightgreen", "pink"),
        main = "Marks Distribution of Subjects",
        ylab = "Marks")
```

```{r}
group1 <- c(150, 152, 155, 158, 160)
group2 <- c(155, 158, 160, 162, 165)
group3 <- c(160, 162, 165, 168, 170)

boxplot(group1, group2, group3,
        names = c("Group A", "Group B", "Group C"),
        col = c("skyblue", "lightgreen", "pink"),
        main = "Height Distribution",
        ylab = "Height (cm)")
```

```{r}
week1 <- c(200, 250, 300, 350, 400)
week2 <- c(300, 350, 400, 450, 500)

boxplot(week1, week2,
        names = c("Week 1", "Week 2"),
        col = c("violet", "gold"),
        main = "Pocket Money Comparison",
        ylab = "Amount (INR)")

grid()
```

#### Histograms

```{r}
marks <- c(45, 50, 55, 60, 62, 65, 68, 70, 72, 75)

hist(marks,
     main = "Histogram of Marks",
     xlab = "Marks")

```

```{r}
hist(marks,
     col = "skyblue",
     border = "yellow",
     main = "Histogram of Marks",
     xlab = "Marks")
```

```{r}
set.seed(123)
data<- rnorm(100)
hist(data,main = "Normal Distribution")

```

```{r}
set.seed(123)

data <- rnorm(100)

hist(data,
     freq = FALSE,
     col = "skyblue",
     border = "darkblue",
     main = "Normal Distribution with Density Curve",
     xlab = "Values")

lines(density(data),
      col = "black",
      lwd = 2)

```

```{r}
plot(marks,
     type = "o",
     col = "blue",
     pch = 16,
     main = "Marks Trend",
     xlab = "Subjects",
     ylab = "Marks")

```

```{r}
study_hours <- c(2, 3, 4, 3, 5)
play_hours  <- c(1, 2, 1, 2, 1)

plot(study_hours,
     type = "o",
     col = "green",
     ylim = c(0, 6),
     xaxt = "n",
     main = "Daily Activity Comparison",
     xlab = "Days",
     ylab = "Hours")

lines(play_hours,
      type = "o",
      col = "blue")

axis(1, at = 1:5, labels = c("Mon", "Tue", "Wed", "Thu", "Fri"))

legend("topleft",
       legend = c("Play", "study"),
       col = c("blue", "green"),
       lty = 1,
       pch = 16)

```

### Data set used

```{r}
#Dataset Load & View
data(mtcars)
View(mtcars)
#command for a data sets 
str(mtcars)
summary(mtcars)
head(mtcars)
```

```{r}
hist(mtcars$mpg,
     col = "skyblue",
     border = "black",
     main = "Mileage Distribution (mpg)",
     xlab = "Miles per Gallon")

```

```{r}
hist(mtcars$mpg,
     freq = FALSE,
     col = "lightgreen",
     border = "black",
     main = "Mileage Distribution with Density Curve",
     xlab = "Miles per Gallon")

lines(density(mtcars$mpg),
      col = "red",
      lwd = 2)

```

```{r}
cyl_count <- table(mtcars$cyl)

barplot(cyl_count,
        col = "orange",
        main = "Number of Cars by Cylinders",
        xlab = "Cylinders",
        ylab = "Count")

```

```{r}
boxplot(mpg ~ cyl,
        data = mtcars,
        col = c("skyblue", "lightgreen", "pink"),
        main = "Mileage vs Cylinders",
        xlab = "Cylinders",
        ylab = "Miles per Gallon")

```

```{r}
plot(mtcars$mpg,
     type = "o",
     col = "blue",
     main = "Mileage of Cars",
     xlab = "Car Index",
     ylab = "Miles per Gallon")

```

```{r}
plot(mtcars$wt, mtcars$mpg,
     col = "red",
     pch = 16,
     main = "Mileage vs Weight",
     xlab = "Weight",
     ylab = "Miles per Gallon")

```
